D-Hammer: Efficient Equational Reasoning for Labelled Dirac Notation

### 2025-01-31 02language_design.tex
18-page version

\begin{enumerate}
    \item \textbf{Term Rewriting by rules}: Expand definitions and simplify expressions.
    \item \textbf{Variable Expansion}: Convert to abstract element-wise representation.
    \item \textbf{Term Rewriting by rules}: Normalize terms on \( R \) modulo \( E \).
    \item \textbf{Sorting Without Bound Variables}: Normalize AC-equivalence.
    \item \textbf{Swap Successive Summations}: Normalize SUM-SWAP equivalence.
    \item \textbf{Use de Bruijn Index}: Normalize \( \alpha \)-equivalence.
\end{enumerate}

>>>>

\begin{enumerate}
    \item \textbf{Rule based term rewriting}: Expand definitions and simplify expressions.
    \item \textbf{Variable expansion}: Convert to abstract element-wise representation.
    \item \textbf{Rule based term rewriting}: Normalize terms on \( R \) modulo \( E \).
    \item \textbf{Sorting without bound variables}: Normalize AC-equivalence.
    \item \textbf{Swapping successive summations}: Normalize SUM-SWAP equivalence.
    \item \textbf{Use de Bruijn index}: Normalize \( \alpha \)-equivalence.
\end{enumerate}

### 2025-01-31 05tool_eval.tex
18-page version

The first three lines use the \texttt{\textcolor{NavyBlue}{Var}} and \texttt{\textcolor{NavyBlue}{Def}} commands to set up the context for the Dirac notation.
\texttt{T} is a type index, representing arbitrary Hilbert space types. \texttt{M} is assume to be an operator in the Hilbert space with type \texttt{T}. \texttt{phi} is defined as the maximally entangled state, depending on the bound variable \texttt{T} as index.
\texttt{r1} and \texttt{r2} are register names for the two subsystems.

@@

\subsubsection{Labelled Dirac Notation}
We present a new set of examples for labelled Dirac notation (LDN). These examples, illustrated in~\Cref{fig: labelled examples}, include six representative cases drawn from a variety of sources, such as well-established theorems, research paper results, quantum circuit equivalence, and quantum program verification.
D-Hammer normalizes them and successfully checks their equivalence using the algorithm in~\Cref{sec: labelled}.
Among the examples, LDN-12 shows the flexibility in combining labelled Dirac notations.
Example LDN-14 and LDN-18 are about circuits or programs on qubits. LDN-14 shows how to calculate controlled-not gate in different ways. In LDN-18, $q, q_1$ and $q_2$ represent variables for quantum programs, and the equation shows the change of the state after applying $X$ gate on $q_2$. 
% Experiments show that our algorithm is relatively inefficient for labelled Dirac notations based on qubits.

>>>>

The first three lines use the \texttt{\textcolor{NavyBlue}{Var}} and \texttt{\textcolor{NavyBlue}{Def}} commands to set up the context for the Dirac notation.
\texttt{T} is a type index, representing arbitrary Hilbert space types. \texttt{M} is assumed to be an operator in the Hilbert space with type \texttt{T}. \texttt{phi} is defined as the maximally entangled state, depending on the bound variable \texttt{T} as index.
\texttt{r1} and \texttt{r2} are register names for the two subsystems.

@@

\subsubsection{Labelled Dirac Notation}
We present a new set of examples for labelled Dirac notation (LDN). These examples, illustrated in~\Cref{fig: labelled examples}, include six representative cases drawn from a variety of sources, such as well-established theorems, research paper results, quantum circuit equivalence, and quantum program verification.
D-Hammer normalizes them and successfully checks their equivalence using the algorithm in~\Cref{sec: labelled}.
Among the examples, LDN-12 shows the flexibility in combining labelled Dirac notations.
LDN-14 shows how to calculate controlled-not gate in different ways. 
%In LDN-18, $q, q_1$ and $q_2$ represent variables for quantum programs, and the equation shows the change of the state after applying $X$ gate on $q_2$. 
% Experiments show that our algorithm is relatively inefficient for labelled Dirac notations based on qubits.

### 2025-01-31 06related_conclusion.tex
18-page version

The most obvious difference is that D-Hammer targets labelled Dirac
notation, whereas DiracDec targets plain Dirac notation. As already
mentioned, the use of labelled Dirac notation is essential in many
applications; in particular, labelled Dirac notation can simplify
notation and proofs, and is in general better suited for writing and
reasoning about complex, many-body, quantum states. However, there are
several other important differences. First, our language leverages the
power of higher-order dependent type theory. In particular, we
leverage higher-order functions to provide a compact and expressive
representation of big operators. Second, our language supports AC
symbols with indefinite arities, which leads to compact representation
of terms, and eases AC reasoning. Third, the relation with Mathematica
is fundamentally different. DiracDec is implemented in Mathematica; as
a consequence, its behavior, and in particular, its typing rules are
constrained by the type system of Mathematica. In contrast, D-Hammer
is designed as a separate tool; as a consequence, it benefits from an
improved representation of terms (e.g.\, AC operators with indefinite
arity), a more expressive type system (e.g.\, dependent types), and a
more efficient rewriting engine. D-Hammer still relies on Mathematica
to reason about functions that are not natively supported by rewriting
rules, but these interactions are constrained and do not have negative
effects on the overall efficiency of the system. This is reflected in
our experimental comparison of D-Hammer and DiracDec, which shows how
the former outperforms the latter.

>>>>

The most obvious difference is that D-Hammer targets labelled Dirac
notation, whereas DiracDec targets plain Dirac notation. As already
mentioned, the use of labelled Dirac notation is essential in many
applications; in particular, labelled Dirac notation can simplify
notation and proofs, and is in general better suited for writing and
reasoning about complex, many-body, quantum states. However, there are
several other important differences. First, our language leverages the
power of higher-order dependent type theory. In particular, we
leverage higher-order functions to provide a compact and expressive
representation of big operators. Second, our language adopts AC
symbols with indefinite arities, which leads to compact representation
of terms, and eases AC reasoning. Third, the relation with Mathematica
is fundamentally different. DiracDec is implemented in Mathematica; as
a consequence, its behavior, and in particular, its typing rules are
constrained by the lack of typing in Mathematica. In contrast, D-Hammer
is designed as a separate tool; as a consequence, it benefits from an
improved representation of terms (e.g.\, AC operators with indefinite
arity), a more expressive type system (e.g.\, dependent types), and a
more efficient rewriting engine. D-Hammer still relies on Mathematica
to reason about functions that are not natively supported by rewriting
rules, but these interactions are constrained and do not have negative
effects on the overall efficiency of the system. This is reflected in
our experimental comparison of D-Hammer and DiracDec, which shows how
the former outperforms the latter.
